---
title: "Project proposal"
author: "RiSK: Rhea Tejwani, Sue Zhang, Keena Gao"
date: "10/6/2020"
output: pdf_document
---


### Section 1. Introduction
The dataset contains statistics of the most popular music in the world over the years on the streaming service Spotify. Music is constantly changing, and as students who frequently listen to music, we want to analyze the trends of popular music in our generation and the generations before us. Our research question is "how have characteristics of the most listened to music changed over time?" Our hypotheses are that popular music has shifted to be more diverse in these categories like acoustics, BPM, valence, etc. 

### Section 2. Data description
This data set has 15 columns and 1994 rows. The observations in the data set describe the characteristics of the top 2000 most popular songs from 1956 to 2019 from Spotify. 
Title: title of the song
Artist: the musician/group who performed the song
Top genre: genre of the track
year: year it was released
Beats per minute(bpm): tempo of the song
Energy: how energetic the song is
Danceability: how easy the song is to dance to
Loudness(dB): how loud the song is
Liveness: the likeliness of the song being a live recording
Valence: how positive a song is
Length(duration): the length of a track
Acousticness: how acoustic a song is
Speechiness: how much spoken word is in the song
Popularity: how popular a song is

The original data set was taken from the playlist on spotify Top 2000s on PlaylistMachinery(@plamere) using Selenium with Python. It was scraped from http://sortyourmusic.playlistmachinery.com/. 

### Section 3. Glimpse of data 
```{r load-data, message=FALSE}
library(tidyverse) 
spotify_data <- read_csv("data/Spotify-2000.csv")
```

```{r glimpse-data}
glimpse(spotify_data)
```


